---
title: "Estimating the Quality of Wines"
subtitle: "With regression trees and model trees"
author: "Shreyas Khadse"
output: md_document
  # prettydoc::html_pretty:
  #     theme: leonids
---

## Introduction 

Winemaking is a challenging and competitive business that offers the potential
for great profit. However, there are numerous factors that contribute to the
profitability of a winery. As an agricultural product, variables as diverse as the
weather and the growing environment impact the quality of a varietal. The bottling
and manufacturing can also affect the flavor for better or worse. Even the way
the product is marketed, from the bottle design to the price point, can affect the
customer's perception of taste.

As a consequence, the winemaking industry has invested heavily in data
collection and machine learning methods that may assist with the decision science
of winemaking.

## Collecting data

To develop the wine rating model, we will use data by P. Cortez, A. Cerdeira,
F. Almeida, T. Matos, and J. Reis. Their dataset includes examples of red and white
Vinho Verde wines from Portugal—one of the world's leading wine-producing
countries. Because the factors that contribute to a highly rated wine may differ
between the red and white varieties, for this analysis we will examine only the
more popular white wines.

The white wine data includes information on 11 chemical properties of 4,898 wine
samples. For each wine, a laboratory analysis measured characteristics such as the
acidity, sugar content, chlorides, sulfur, alcohol, pH, and density. The samples were
then rated in a blind tasting by panels of no less than three judges on a quality scale
ranging from zero (very bad) to 10 (excellent). In the case that the judges disagreed
on the rating, the median value was used.

For ease to acces the dataset I have hosted a public repository and included the code 
that directly downloads the data from the repo and loads the data. The path can be canged 
to anythig according to your preference of the working directory.

```{r}
set.seed(6799)
path <- "A:/Project/Quality_Wines"
setwd(path)
url <- "https://raw.githubusercontent.com/shreyaskhadse/data_files/master/whitewines.csv"
datafile <- "./whitewines.csv"
if (!file.exists(datafile)) {
    download.file(url, datafile ,method="auto") }
wine <- read.csv("whitewines.csv")
str(wine)
```


## Exploring and Cleaning data

Compared with other types of machine learning models, one of the advantages of
trees is that they can handle many types of data without preprocessing. This means
we do not need to normalize or standardize the features.

We now examine the distribution of the outcome variable is needed
to inform our evaluation of the model's performance. For instance, suppose that
there was very little variation in quality from wine to wine, or that wines fell into
a bimodal distribution: either very good or very bad. This may impact the way we
design the model. To check for such extremes, we can examine the distribution of
wine quality using a histogram:

```{r}
library(ggplot2)
ggplot( wine, aes(x = quality))+geom_histogram(aes(y = ..density..), color = "black", binwidth = 1)+labs(title = "Histogram of wine quality", x = "Qualtiy", y = "Frequency") +
  stat_function(fun = dnorm, args = list(mean = mean(wine$quality),sd = sd(wine$quality)),col = "steelblue",size = 1)
```

The wine quality values appear to follow a fairly normal, bell-shaped distribution,
centered around a value of six. This makes sense intuitively, because most wines
are of average quality; few are particularly bad or good.

Now we divide the dataset into training and testing sets. Since
the wine dataset was already sorted randomly, we can partition into two sets
of contiguous rows as follows:

```{r}
wine_train <- wine[1:3750,]
wine_test <- wine[3751:4898,]
```

## Training a model on the data

We will begin by training a regression tree model. Although almost any
implementation of decision trees can be used to perform regression tree modeling,
the `rpart` (recursive partitioning) package offers the most faithful implementation
of regression trees as they were described by the CART team.

```{r}
#install.packages("rpart")
library(rpart)
m.rpart <- rpart(quality ~ ., data = wine_train)
m.rpart
summary(m.rpart)
```

`alcohol` has the highest importance as a predictor, followed by `density`.

For each node in the tree, the number of examples reaching the decision point is
listed. For instance, all 3,750 examples begin at the root node, of which 2,473 have
`alcohol < 10.85` and 1,277 have `alcohol >= 10.85`. Because alcohol was used
first in the tree, it is the single most important predictor of wine quality.

Nodes indicated by * are terminal or leaf nodes, which means that they result
in a prediction (listed here as yval). For example, node 5 has a `yval` of 5.881912.
When the tree is used for predictions, any wine samples with `alcohol < 10.85`
and volatile.acidity < 0.2425 would therefore be predicted to have a quality
value of 5.881912.

```{r}
#install.packages("rpart.plot")
library(rpart.plot)
rpart.plot(m.rpart, digits = 3, box.palette = "BuGn")
#rpart.plot(m.rpart, digits = 4, fallen.leaves = TRUE,type = 3, extra = 101)
```


## Evaluating model performance

To use the regression tree model to make predictions on the test data, we use the
`predict()` function. By default, this returns the estimated numeric value for the
outcome variable, which we'll save in a vector named `p.rpart`:

```{r}
p.rpart <- predict(m.rpart, wine_test)
"summary(p.part)"
summary(p.rpart)

"summary(wine_test$quality)"
summary(wine_test$quality)
```

This finding suggests that the model is not correctly identifying the extreme cases,
in particular, the best and worst wines. On the other hand, between the first and
third quartile, we may be doing well.

The correlation between the predicted and actual quality values provides a simple
way to gauge the model's performance.

```{r}
cor(p.rpart, wine_test$quality)
```

A correlation of 0.49 is certainly acceptable. However, the correlation only measures
how strongly the predictions are related to the true value; it is not a measure of how
far off the predictions were from the true values.

Another way to think about the model's performance is to consider how far, on
average, its prediction was from the true value. This measurement is called the
mean absolute error (MAE).

```{r}
MAE <- function(actual, predicted) {mean(abs(actual - predicted))}
MAE(p.rpart, wine_test$quality)
```

This implies that, on average, the difference between our model's predictions and
the true quality score was about 0.59. On a quality scale from zero to 10, this seems
to suggest that our model is doing fairly well.

On the other hand, recall that most wines were neither very good nor very bad; the
typical quality score was around five to six. Therefore, a classifier that did nothing
but predict the mean value may still do fairly well according to this metric.

The mean quality rating in the training data is as follows:

```{r}
mean(wine_train$quality)
```

If we predicted the value 5.89 for every wine sample, we would have a mean
absolute error of only about 0.58:

```{r}
MAE(5.87, wine_test$quality)
```

Our regression tree (MAE = 0.57) comes closer on average to the true quality score
than the imputed mean (MAE = 0.58), but not by much. In comparison, Cortez
reported an MAE of 0.58 for the neural network model and an MAE of 0.45 for
the support vector machine. This suggests that there is room for improvement.

## Improving model performance

To improve the performance of our learner, let's apply a model tree algorithm, which
is a more complex application of trees to numeric prediction.A model tree
extends regression trees by replacing the leaf nodes with regression models. This
often results in more accurate results than regression trees, which use only a single
numeric value for the prediction at the leaf nodes.

The current state-of-the-art in model trees is the Cubist algorithm, which itself is
an enhancement of the M5 model tree algorithm. The Cubist algorithm involves building a decision tree, creating decision rules based on the branches of the tree, and building a regression model
at each of the leaf nodes. Additional heuristics, such as pruning and boosting, are
used to improve the quality of the predictions and smoothness across the range of
predicted values.

```{r}
#install.packages("Cubist")
library(Cubist)
m.cubist <- cubist(x = wine_train[-12], y = wine_train$quality)
m.cubist
summary(m.cubist)
```

In this output, we see that the algorithm generated 10 rules to model the wine
quality.

You will note that the `if` portion of the output is somewhat similar to the regression
tree we built earlier. A series of decisions based on the wine properties of sulfur
dioxide, sulphates, and alcohol creates a rule culminating in the final prediction.
A key difference between this model tree output and the earlier regression tree
output, however, is that the nodes here terminate not in a numeric prediction,
but rather a linear model.

The linear model for this rule is shown in the then output following the `outcome = statement`

To examine the performance of this model, we'll look at how well it performs on
the unseen test data. The predict() function gets us a vector of predicted values:

```{r}
p.cubist <- predict(m.cubist, wine_test)
summary(p.cubist)
```
The correlation also seems to be substantially higher:

```{r}
"cor()"
cor(p.cubist, wine_test$quality)
"MAE"
MAE(wine_test$quality, p.cubist)
```

Although we did not improve a great deal beyond the regression tree, we surpassed
the performance of the neural network model published by Cortez, and we are
getting closer to the published mean absolute error value of 0.45 for the support
vector machine model, all while using a much simpler learning method.














