---
title: "HW02 - Filtering mobile phone spam with the Naive Bayes algorithm"
author: "Hoa Quach"
output: html_notebook
---

#Filtering mobile phone spam with the Naive Bayes algorithm

As the worldwide use of mobile phones has grown, a new avenue for electronic
junk mail has opened for disreputable marketers. These advertisers utilize Short
Message Service (SMS) text messages to target potential consumers with unwanted
advertising known as SMS spam. This type of spam is particularly troublesome
because, unlike e-mail spam, many cellular phone users pay a fee per SMS received.
Developing a classification algorithm that could filter SMS spam would provide a
useful tool for cellular phone providers.

Since Naive Bayes has been used successfully for e-mail spam filtering, it seems
likely that it could also be applied to SMS spam. However, relative to e-mail spam,
SMS spam poses additional challenges for automated filters. SMS messages are often
limited to 160 characters, reducing the amount of text that can be used to identify
whether a message is junk. The limit, combined with small mobile phone keyboards,
has led many to adopt a form of SMS shorthand lingo, which further blurs the line
between legitimate messages and spam. Let's see how a simple Naive Bayes classifier
handles these challenges.

#Data Collection

To develop the Naive Bayes classifier, we will use data adapted from the SMS Spam
Collection at http://www.dt.fee.unicamp.br/~tiago/smsspamcollection/.

This dataset includes the text of SMS messages along with a label indicating
whether the message is unwanted. Junk messages are labeled spam, while
legitimate messages are labeled ham. Some examples of spam and ham are
shown in the following table:

##Sample SMS ham

* Better. Made up for Friday and
stuffed myself like a pig yesterday.
Now I feel bleh. But, at least, its not
writhing pain kind of bleh.
* If he started searching, he will get job
in few days. He has great potential
and talent.
* I got another job! The one at the
hospital, doing data analysis or
something, starts on Monday! Not
sure when my thesis will finish.

##Sample SMS spam

* Congratulations ur awarded 500 of
CD vouchers or 125 gift guaranteed
& Free entry 2 100 wkly draw txt
MUSIC to 87066.
* December only! Had your mobile
11mths+? You are entitled to update
to the latest colour camera mobile
for Free! Call The Mobile Update Co
FREE on 08002986906.
* Valentines Day Special! Win over
£1000 in our quiz and take your
partner on the trip of a lifetime! Send
GO to 83600 now. 150 p/msg rcvd.

Looking at the preceding messages, did you notice any distinguishing characteristics
of spam? One notable characteristic is that two of the three spam messages use the
word "free," yet the word does not appear in any of the ham messages. On the other
hand, two of the ham messages cite specific days of the week, as compared to zero in
spam messages.

Our Naive Bayes classifier will take advantage of such patterns in the word
frequency to determine whether the SMS messages seem to better fit the profile of
spam or ham. While it's not inconceivable that the word "free" would appear outside
of a spam SMS, a legitimate message is likely to provide additional words explaining
the context. For instance, a ham message might state "are you free on Sunday?"
Whereas, a spam message might use the phrase "free ringtones." The classifier will
compute the probability of spam and ham, given the evidence provided by all the
words in the message.


#Data Exploration

The first step towards constructing our classifier involves processing the raw data for
analysis. Text data are challenging to prepare, because it is necessary to transform the
words and sentences into a form that a computer can understand. We will transform
our data into a representation known as bag-of-words, which ignores word order and
simply provides a variable indicating whether the word appears at all.

We'll begin by importing the CSV data and saving it in a data frame:
```{r}
sms_raw <- read.csv("sms_spam.csv", stringsAsFactors = FALSE)
```

Using the str() function, we see that the sms_raw data frame includes 5,559 total
SMS messages with two features: type and text. The SMS type has been coded as
either ham or spam. The text element stores the full raw SMS text.
```{r}
str(sms_raw)
```

The type element is currently a character vector. Since this is a categorical variable, it
would be better to convert it into a factor, as shown in the following code:

```{r}
sms_raw$type <- factor(sms_raw$type)
```

Examining this with the str() and table() functions, we see that type has now
been appropriately recoded as a factor. Additionally, we see that 747 (about 13
percent) of SMS messages in our data were labeled as spam, while the others were
labeled as ham:

```{r}
str(sms_raw$type)
```

```{r}
table(sms_raw$type)
```

For now, we will leave the message text alone. As you will learn in the next section,
processing the raw SMS messages will require the use of a new set of powerful tools
designed specifically to process text data.

##Data preparation - cleaning and standardizing text data

SMS messages are strings of text composed of words, spaces, numbers, and
punctuation. Handling this type of complex data takes a lot of thought and
effort. One needs to consider how to remove numbers and punctuation; handle
uninteresting words such as and, but, and or; and how to break apart sentences into
individual words. Thankfully, this functionality has been provided by the members
of the R community in a text mining package titled tm.

The tm package can be installed via the install.packages("tm") command
and loaded with the library(tm) command. Even if you already have it installed,
it may be worth re-running the install process to ensure that your version is
up-to-date, as the tm package is still being actively developed. This occasionally
results in changes to its functionality.

The first step in processing text data involves creating a corpus, which is a collection
of text documents. The documents can be short or long, from individual news
articles, pages in a book or on the web, or entire books. In our case, the corpus
will be a collection of SMS messages.

In order to create a corpus, we'll use the VCorpus() function in the tm package,
which refers to a volatile corpus-volatile as it is stored in memory as opposed to
being stored on disk (the PCorpus() function can be used to access a permanent
corpus stored in a database). This function requires us to specify the source of
documents for the corpus, which could be from a computer's filesystem, a database,
the Web, or elsewhere. Since we already loaded the SMS message text into R, we'll
use the VectorSource() reader function to create a source object from the existing
sms_raw$text vector, which can then be supplied to VCorpus() as follows:


```{r}
library(tm)
sms_corpus <- VCorpus(VectorSource(sms_raw$text))
```

The resulting corpus object is saved with the name sms_corpus.

By printing the corpus, we see that it contains documents for each of the 5,559 SMS
messages in the training data:

```{r}
print(sms_corpus)
```

Because the tm corpus is essentially a complex list, we can use list operations to select
documents in the corpus. To receive a summary of specific messages, we can use the
inspect() function with list operators. For example, the following command will
view a summary of the first and second SMS messages in the corpus:

```{r}
inspect(sms_corpus[1:2])
```

To view the actual message text, the as.character() function must be applied to
the desired messages. To view one message, use the as.character() function on
a single list element, noting that the double-bracket notation is required:

```{r}
as.character(sms_corpus[[1]])
```

To view multiple documents, we'll need to use as.character() on several items in
the sms_corpus object. To do so, we'll use the lapply() function, which is a part of a
family of R functions that applies a procedure to each element of an R data structure.
These functions, which include apply() and sapply() among others, are one of the
key idioms of the R language. Experienced R coders use these much like the way
for or while loops are used in other programming languages, as they result in more
readable (and sometimes more efficient) code. The lapply() command to apply
as.character() to a subset of corpus elements is as follows:

```{r}
lapply(sms_corpus[1:2], as.character)
```

As noted earlier, the corpus contains the raw text of 5,559 text messages. In order
to perform our analysis, we need to divide these messages into individual words.
But first, we need to clean the text, in order to standardize the words, by removing
punctuation and other characters that clutter the result. For example, we would like
the strings Hello!, HELLO, and hello to be counted as instances of the same word.

The tm_map() function provides a method to apply a transformation (also known
as mapping) to a tm corpus. We will use this function to clean up our corpus using a
series of transformations and save the result in a new object called corpus_clean.

Our first order of business will be to standardize the messages to use only lowercase
characters. To this end, R provides a tolower() function that returns a lowercase
version of text strings. In order to apply this function to the corpus, we need to
use the tm wrapper function content_transformer() to treat tolower() as a
transformation function that can be used to access the corpus. The full command is
as follows:

```{r}
sms_corpus_clean <- tm_map(sms_corpus, content_transformer(tolower))
```

To check whether the command worked as advertised, let's inspect the first message
in the original corpus and compare it to the same in the transformed corpus:

```{r}
as.character(sms_corpus[[1]])
```

```{r}
as.character(sms_corpus_clean[[1]])
```

As expected, uppercase letters have been replaced by lowercase versions of the same.

Let's continue our cleanup by removing numbers from the SMS messages. Although
some numbers may provide useful information, the majority would likely be unique
to individual senders and thus will not provide useful patterns across all messages.
With this in mind, we'll strip all the numbers from the corpus as follows:

```{r}
sms_corpus_clean <- tm_map(sms_corpus_clean, removeNumbers)
```

Our next task is to remove filler words such as to, and, but, and or from our SMS
messages. These terms are known as stop words and are typically removed prior to
text mining. This is due to the fact that although they appear very frequently, they do
not provide much useful information for machine learning.

Rather than define a list of stop words ourselves, we'll use the stopwords() function
provided by the tm package. This function allows us to access various sets of stop
words, across several languages. By default, common English language stop words
are used. To see the default list, type stopwords() at the command line. To see the
other languages and options available, type ?stopwords for the documentation page.

The stop words alone are not a useful transformation. What we need is a way
to remove any words that appear in the stop words list. The solution lies in the
removeWords() function, which is a transformation included with the tm package.
As we have done before, we'll use the tm_map() function to apply this mapping to
the data, providing the stopwords() function as a parameter to indicate exactly the
words we would like to remove. The full command is as follows:

```{r}
sms_corpus_clean <- tm_map(sms_corpus_clean, removeWords, stopwords())
```

Since stopwords() simply returns a vector of stop words, had we chosen so, we
could have replaced it with our own vector of words to be removed. In this way, we
could expand or reduce the list of stop words to our liking or remove a completely
different set of words entirely.

Continuing with our cleanup process, we can also eliminate any punctuation from
the text messages using the built-in removePunctuation() transformation:

```{r}
sms_corpus_clean <- tm_map(sms_corpus_clean, removePunctuation)
```

The removePunctuation() transformation strips punctuation characters from the
text blindly, which can lead to unintended consequences. For example, consider
what happens when it is applied as follows:

```{r}
removePunctuation("hello...world")
```

As shown, the lack of blank space after the ellipses has caused the words hello and
world to be joined as a single word. While this is not a substantial problem for our
analysis, it is worth noting for the future.

Another common standardization for text data involves reducing words to their root
form in a process called stemming. The stemming process takes words like learned,
learning, and learns, and strips the suffix in order to transform them into the base
form, learn. This allows machine learning algorithms to treat the related terms as a
single concept rather than attempting to learn a pattern for each variant.

The tm package provides stemming functionality via integration with the SnowballC
package. At the time of this writing, SnowballC was not installed by default with tm.
Do so with install.packages("SnowballC") if it is not installed already.

The SnowballC package provides a wordStem() function, which for a character
vector, returns the same vector of terms in its root form. For example, the function
correctly stems the variants of the word learn, as described previously:

```{r}
library(SnowballC)
```

In order to apply the wordStem() function to an entire corpus of text documents, the
tm package includes a stemDocument() transformation. We apply this to our corpus
with the tm_map() function exactly as done earlier:

```{r}
sms_corpus_clean <- tm_map(sms_corpus_clean, stemDocument)
```

After removing numbers, stop words, and punctuation as well as performing
stemming, the text messages are left with the blank spaces that previously separated
the now-missing pieces. The final step in our text cleanup process is to remove
additional whitespace, using the built-in stripWhitespace() transformation:

```{r}
sms_corpus_clean <- tm_map(sms_corpus_clean, stripWhitespace)
```

The following table shows the first three messages in the SMS corpus before and
after the cleaning process. The messages have been limited to the most interesting
words, and punctuation and capitalization have been removed:

##SMS messages before cleaning

```{r}
as.character(sms_corpus[1:3])
```

##SMS messages after cleaning

```{r}
as.character(sms_corpus_clean[1:3])
```

##Data preparation - splitting text documents into words

Now that the data are processed to our liking, the final step is to split the messages
into individual components through a process called tokenization. A token is a
single element of a text string; in this case, the tokens are words.

As you might assume, the tm package provides functionality to tokenize the SMS
message corpus. The DocumentTermMatrix() function will take a corpus and create
a data structure called a Document Term Matrix (DTM) in which rows indicate
documents (SMS messages) and columns indicate terms (words).

Creating a DTM sparse matrix, given a tm corpus, involves a single command:
```{r}
sms_dtm <- DocumentTermMatrix(sms_corpus_clean)
```

This will create an sms_dtm object that contains the tokenized corpus using
the default settings, which apply minimal processing. The default settings
are appropriate because we have already prepared the corpus manually.

On the other hand, if we hadn't performed the preprocessing, we could do so
here by providing a list of control parameter options to override the defaults.
For example, to create a DTM directly from the raw, unprocessed SMS corpus,
we can use the following command:

```{r}
sms_dtm2 <- DocumentTermMatrix(sms_corpus, control = list(
  tolower = TRUE,
  removeNumbers = TRUE, 
  stopwords = TRUE, 
  removePunctuation = TRUE, 
  stemming = TRUE))
```

This applies the same preprocessing steps to the SMS corpus in the same order as
done earlier. However, comparing sms_dtm to sms_dtm2, we see a slight difference
in the number of terms in the matrix:

```{r}
sms_dtm
```

```{r}
sms_dtm2
```

The reason for this discrepancy has to do with a minor difference in the ordering of
the preprocessing steps. The DocumentTermMatrix() function applies its cleanup
functions to the text strings only after they have been split apart into words. Thus,
it uses a slightly different stop words removal function. Consequently, some words
split differently than when they are cleaned before tokenization.

The differences between these two cases illustrate an important principle of cleaning
text data: the order of operations matters. With this in mind, it is very important
to think through how early steps in the process are going to affect later ones. The
order presented here will work in many cases, but when the process is tailored more
carefully to specific datasets and use cases, it may require rethinking. For example,
if there are certain terms you hope to exclude from the matrix, consider whether you
should search for them before or after stemming. Also, consider how the removal
of punctuation-and whether the punctuation is eliminated or replaced by blank
space-affects these steps.

# Data preparation
## creating training and test datasets

With our data prepared for analysis, we now need to split the data into training and
test datasets, so that once our spam classifier is built, it can be evaluated on data it
has not previously seen. But even though we need to keep the classifier blinded as to
the contents of the test dataset, it is important that the split occurs after the data have
been cleaned and processed; we need exactly the same preparation steps to occur on
both the training and test datasets.

We'll divide the data into two portions: 75 percent for training and 25 percent for
testing. Since the SMS messages are sorted in a random order, we can simply take the
first 4,169 for training and leave the remaining 1,390 for testing. Thankfully, the DTM
object acts very much like a data frame and can be split using the standard [row,
col] operations. As our DTM stores SMS messages as rows and words as columns,
we must request a specific range of rows and all columns for each:

```{r}
sms_dtm_train <- sms_dtm[1:4169, ]
sms_dtm_test <- sms_dtm[4170:5559, ]
```

For convenience later on, it is also helpful to save a pair of vectors with labels for
each of the rows in the training and testing matrices. These labels are not stored in
the DTM, so we would need to pull them from the original sms_raw data frame:

```{r}
sms_train_labels <- sms_raw[1:4169, ]$type
sms_test_labels <- sms_raw[4170:5559, ]$type
```

To confirm that the subsets are representative of the complete set of SMS data, let's
compare the proportion of spam in the training and test data frames:

```{r}
prop.table(table(sms_train_labels))
```

```{r}
prop.table(table(sms_test_labels))
```

Both the training data and test data contain about 13 percent spam. This suggests
that the spam messages were divided evenly between the two datasets.

## Visualizing text data - word clouds

A word cloud is a way to visually depict the frequency at which words appear in
text data. The cloud is composed of words scattered somewhat randomly around the
figure. Words appearing more often in the text are shown in a larger font, while less
common terms are shown in smaller fonts. This type of figures grew in popularity
recently, since it provides a way to observe trending topics on social media websites.

The wordcloud package provides a simple R function to create this type of diagrams.
We'll use it to visualize the types of words in SMS messages, as comparing the clouds
for spam and ham will help us gauge whether our Naive Bayes spam filter is likely to
be successful. If you haven't already done so, install and load the package by typing
install.packages("wordcloud") and library(wordcloud) at the R command line.

A word cloud can be created directly from a tm corpus object using the syntax and the resulting word cloud should appear similar to the following figure. 

This will create a word cloud from our prepared SMS corpus. Since we specified
random.order = FALSE, the cloud will be arranged in a nonrandom order with higher
frequency words placed closer to the center. If we do not specify random.order, the
cloud would be arranged randomly by default. The min.freq parameter specifies the
number of times a word must appear in the corpus before it will be displayed in the
cloud. Since a frequency of 50 is about 1 percent of the corpus, this means that a word
must be found in at least 1 percent of the SMS messages to be included in the cloud.

```{r}
library(wordcloud)
wordcloud(sms_corpus_clean, min.freq = 50, random.order = FALSE)
```

A perhaps more interesting visualization involves comparing the clouds for SMS
spam and ham. Since we did not construct separate corpora for spam and ham,
this is an appropriate time to note a very helpful feature of the wordcloud()
function. Given a vector of raw text strings, it will automatically apply common text
preparation processes before displaying the cloud.

Let's use R's subset() function to take a subset of the sms_raw data by the SMS
type. First, we'll create a subset where the message type is spam:

```{r}
spam <- subset(sms_raw, type == "spam")
```

Next, we'll do the same thing for the ham subset:

```{r}
ham <- subset(sms_raw, type == "ham")
```

We now have two data frames, spam and ham, each with a text feature containing
the raw text strings for SMSes. Creating word clouds is as simple as before. This
time, we'll use the max.words parameter to look at the 40 most common words in
each of the two sets. The scale parameter allows us to adjust the maximum and
minimum font size for words in the cloud. Feel free to adjust these parameters as
you see fit. This is illustrated in the following commands and the resulting word clouds are shown in the following diagrams:

```{r}
wordcloud(spam$text, max.words = 40, scale = c(3, 0.5))
wordcloud(ham$text, max.words = 40, scale = c(3, 0.5))
```

```{r}
```

Do you have a hunch about which one is the spam cloud and which represents ham?

As you probably guessed, the spam cloud is on the left. Spam messages include
words such as urgent, free, mobile, claim, and stop; these terms do not appear in the
ham cloud at all. Instead, ham messages use words such as can, sorry, need, and time.
These stark differences suggest that our Naive Bayes model will have some strong
key words to differentiate between the classes.

# Data preparation
## creating indicator features for frequent words

The final step in the data preparation process is to transform the sparse matrix into a
data structure that can be used to train a Naive Bayes classifier. Currently, the sparse
matrix includes over 6,500 features; this is a feature for every word that appears in at
least one SMS message. It's unlikely that all of these are useful for classification. To
reduce the number of features, we will eliminate any word that appear in less than five
SMS messages, or in less than about 0.1 percent of the records in the training data.

Finding frequent words requires use of the findFreqTerms() function in the
tm package. This function takes a DTM and returns a character vector containing
the words that appear for at least the specified number of times. For instance,
the following command will display the words appearing at least five times in
the sms_dtm_train matrix:

```{r}
findFreqTerms(sms_dtm_train, 5)
```

The result of the function is a character vector, so let's save our frequent words for
later on:

```{r}
sms_freq_words <- findFreqTerms(sms_dtm_train, 5)
```

A peek into the contents of the vector shows us that there are 1,136 terms appearing
in at least five SMS messages:

```{r}
str(sms_freq_words)
```

We now need to filter our DTM to include only the terms appearing in a specified
vector. As done earlier, we'll use the data frame style [row, col] operations to
request specific portions of the DTM, noting that the columns are named after the
words the DTM contains. We can take advantage of this to limit the DTM to specific
words. Since we want all the rows, but only the columns representing the words in
the sms_freq_words vector, our commands are:

```{r}
sms_dtm_freq_train<- sms_dtm_train[ , sms_freq_words]
sms_dtm_freq_test <- sms_dtm_test[ , sms_freq_words]
```

The training and test datasets now include 1,136 features, which correspond to
words appearing in at least five messages.

The Naive Bayes classifier is typically trained on data with categorical features.
This poses a problem, since the cells in the sparse matrix are numeric and measure
the number of times a word appears in a message. We need to change this to a
categorical variable that simply indicates yes or no depending on whether the
word appears at all.

The following defines a convert_counts() function to convert counts to
Yes/No strings:

```{r}
convert_counts <- function(x) {
  x <- ifelse(x > 0, "Yes", "No")
}
```

By now, some of the pieces of the preceding function should look familiar. The first
line defines the function. The ifelse(x > 0, "Yes", "No") statement transforms
the values in x, so that if the value is greater than 0, then it will be replaced by "Yes",
otherwise it will be replaced by a "No" string. Lastly, the newly transformed x vector
is returned.

We now need to apply convert_counts() to each of the columns in our sparse
matrix. You may be able to guess the R function to do exactly this. The function
is simply called apply() and is used much like lapply() was used previously.

The apply() function allows a function to be used on each of the rows or columns
in a matrix. It uses a MARGIN parameter to specify either rows or columns. Here,
we'll use MARGIN = 2, since we're interested in the columns (MARGIN = 1 is used
for rows). The commands to convert the training and test matrices are as follows:

```{r}
sms_train <- apply(sms_dtm_freq_train, MARGIN = 2,
convert_counts)
```

```{r}
sms_test <- apply(sms_dtm_freq_test, MARGIN = 2,
convert_counts)
```

The result will be two character type matrixes, each with cells indicating "Yes" or
"No" for whether the word represented by the column appears at any point in the
message represented by the row.

# Step 3 - training a model on the data

Now that we have transformed the raw SMS messages into a format that can be
represented by a statistical model, it is time to apply the Naive Bayes algorithm. The
algorithm will use the presence or absence of words to estimate the probability that a
given SMS message is spam.

The Naive Bayes implementation we will employ is in the e1071 package. This
package was developed in the statistics department of the Vienna University of
Technology (TU Wien), and includes a variety of functions for machine learning.
If you have not done so already, be sure to install and load the package using the
install.packages("e1071") and library(e1071) commands before continuing.

Unlike the k-NN algorithm we used for classification in the previous chapter, a
Naive Bayes learner is trained and used for classification in separate stages. Still,
as shown in the following table, these steps are is fairly straightforward.

To build our model on the sms_train matrix, we'll use the following command:

```{r}
library(e1071)
library(klaR)
sms_classifier <- naiveBayes(sms_train, sms_train_labels)
```


The sms_classifier object now contains a naiveBayes classifier object that can be
used to make predictions.

# Step 4 - evaluating model performance

To evaluate the SMS classifier, we need to test its predictions on unseen messages
in the test data. Recall that the unseen message features are stored in a matrix
named sms_test, while the class labels (spam or ham) are stored in a vector named
sms_test_labels. The classifier that we trained has been named sms_classifier.
We will use this classifier to generate predictions and then compare the predicted
values to the true values.

The predict() function is used to make the predictions. We will store these in a
vector named sms_test_pred. We will simply supply the function with the names
of our classifier and test dataset, as shown:

```{r}
sms_test_pred <- predict(sms_classifier, sms_test)
```

To compare the predictions to the true values, we'll use the CrossTable() function
in the gmodels package, which we used previously. This time, we'll add some
additional parameters to eliminate unnecessary cell proportions and use the dnn
parameter (dimension names) to relabel the rows and columns, as shown in the
following code:

```{r}
library(gmodels)
CrossTable(sms_test_pred, sms_test_labels,
prop.chisq = FALSE, prop.t = FALSE,
dnn = c('predicted', 'actual'))

```

This produces the following table:

Looking at the table, we can see that a total of only 6 + 30 = 36 of the 1,390 SMS
messages were incorrectly classified (2.6 percent). Among the errors were 6 out
of 1,207 ham messages that were misidentified as spam, and 30 of the 183 spam
messages were incorrectly labeled as ham. Considering the little effort we put
into the project, this level of performance seems quite impressive. This case study
exemplifies the reason why Naive Bayes is the standard for text classification;
directly out of the box, it performs surprisingly well.

On the other hand, the six legitimate messages that were incorrectly classified
as spam could cause significant problems for the deployment of our filtering
algorithm, because the filter could cause a person to miss an important text message.
We should investigate to see whether we can slightly tweak the model to achieve
better performance.

# Step 5 - improving model performance

You may have noticed that we didn't set a value for the Laplace estimator while
training our model. This allows words that appeared in zero spam or zero ham
messages to have an indisputable say in the classification process. Just because the
word "ringtone" only appeared in the spam messages in the training data, it does
not mean that every message with this word should be classified as spam.

We'll build a Naive Bayes model as done earlier, but this time set laplace = 1:

```{r}
sms_classifier2 <- naiveBayes(sms_train, sms_train_labels,
laplace = 1)
```

Next, we'll make predictions:

```{r}
sms_test_pred2 <- predict(sms_classifier2, sms_test)
```

Finally, we'll compare the predicted classes to the actual classifications using a
cross tabulation and the results are shown below:

```{r}
CrossTable(sms_test_pred2, sms_test_labels,
prop.chisq = FALSE, prop.t = FALSE, prop.r = FALSE,
dnn = c('predicted', 'actual'))
```

Adding the Laplace estimator reduced the number of false positives (ham messages
erroneously classified as spam) from six to five and the number of false negatives
from 30 to 28. Although this seems like a small change, it's substantial considering
that the model's accuracy was already quite impressive. We'd need to be careful
before tweaking the model too much in order to maintain the balance between being
overly aggressive and overly passive while filtering spam. Users would prefer that a
small number of spam messages slip through the filter than an alternative in which
ham messages are filtered too aggressively.

# Summary

In this chapter, we learned about classification using Naive Bayes. This algorithm
constructs tables of probabilities that are used to estimate the likelihood that new
examples belong to various classes. The probabilities are calculated using a formula
known as Bayes' theorem, which specifies how dependent events are related.
Although Bayes' theorem can be computationally expensive, a simplified version that
makes so-called "naive" assumptions about the independence of features is capable
of handling extremely large datasets.

The Naive Bayes classifier is often used for text classification. To illustrate its
effectiveness, we employed Naive Bayes on a classification task involving spam
SMS messages. Preparing the text data for analysis required the use of specialized
R packages for text processing and visualization. Ultimately, the model was able to
classify over 97 percent of all the SMS messages correctly as spam or ham.


###EOF

### References
Lantz, Brett. Machine Learning with R. 2nd ed. Birmingham: Packt Publishing Ltd, 2015. Print. , 2013. Print.